# The WDC (World Drivers Championship)

Throughout a season, teams and drivers compete for the WCC (World Constructor Championship) and WDC (World Drivers Championship), respectively. The WCC is awarded to the team with the highest total points scored and the WDC is awarded to the driver with the highest total points. During some seasons, the champions are crowned early (i.e. 2023 and 2022), while other championships may not be decided until the final lap of the season (i.e. 2021). 

```{r, include = F, message = F, warning = F}
library(tidyverse)
library(broom)
library(ggtext)
library(ciTools)
library(merTools)
library(ggrepel)
library(emmeans)
library(gghighlight)
library(ggtext)
library(gt)
library(drs)
library(ggh4x)



# Create title color coder function
title_color_coder <- function(leading_regular_text, label_1, color_1, middle_regular_text, label_2, color_2, trailing_regular_text){

  title_text <- paste0(leading_regular_text, "<span style='color:", color_1,";'>",label_1,"</span>", middle_regular_text, "<span style='color:", color_2,";'>", label_2, "</span>", trailing_regular_text)

  return(title_text)

}



```


Throughout this chapter, I will plot the cumulative points scored for each driver, working backward chronologically. 

In 2023, we saw one of the most dominant seasons ever by a driver. Max won 19 of 22 races and finished on the podium in every race except Singapore. Lewis Hamilton and Sergio Perez fought for second place in the WDC. Additionally, Fernando Alonso made a welcome return to the front of the grid! 

```{r, warning = F, message = F, fig.align='center', fig.height= 10, fig.width=12}


# Pull race results
races2023 <- race_result_scraper(2023)

races2023 %>%
  mutate(rowid = 1:n()) %>%
  group_by(Driver) %>% 
  mutate(cumulativePoints = cumsum(Points)) %>%
  ungroup() %>% 
  ggplot(aes(fct_reorder(Race, rowid), cumulativePoints,
         group = Driver, col = Driver)) +
  geom_line() +
  theme_bw() + 
  theme(
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12),
        axis.text.y = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 18),
        legend.text = element_text(size = 12),
        legend.title = element_text(hjust = 0.5, size = 14)) +
  labs(x = 'Grand Prix',
       y = 'Cumulative Points',
       title = '2023 Formula 1 WDC Race') +
  scale_color_manual(values = c("red", "red", 
                                "cyan3", "cyan3", 
                                "black", "black",
                                "blue", "blue",
                                "darkslateblue", "darkslateblue",
                                "darkslateblue", "darkslateblue",
                                "darkred", "darkred", 
                                "darkseagreen4", "darkseagreen4",
                                "darkorange", "darkorange",
                                "cornflowerblue", "cornflowerblue",
                                "navy", "navy"),
                     breaks = c("LEC", "SAI",
                                "HAM", "RUS", 
                                "MAG", "HUL", 
                                "OCO", "GAS",
                                "TSU",  "DEV", "RIC", "LAW",
                                "ZHO", "BOT",
                                "STR", "ALO", 
                                "PIA", "NOR",
                                "SAR", "ALB",
                                "PER", "VER"
                                )) +
  geom_text_repel(data = ~ subset(.x, Driver %in% c('VER', 'PER', 'HAM', 'ALO') & Race == 'abu-dhabi'),
            aes(label = Driver),
            nudge_x = 2, nudge_y = 3, show.legend = F)
```

<br>
<br>

:::: {.blackbox data-latex=""}
::: {.center data-latex=""}
**Labeling the end of a line in ggplot**
:::

In the plot above, I used the `geom_text_repel()` function from the **ggrepel** package to label the top four point scoring drivers in the 2023 WDC. **ggrepel** provides geometries for ggplot2 to *repel* overlapping text labels. These two geometries are:

* `geom_text_repel()`
* `geom_label_repel()`

These functions cause the text labels to repel away from each other, away from data points, and away from edges of the plotting area (panel).

For more information on this package, visit:

https://cran.r-project.org/web/packages/ggrepel/vignettes/ggrepel.html

***Note:*** In the figure above, I *subset* the data being fed to `geom_text_repel()`. In this example, I include four drivers ('VER', 'PER', 'HAM', 'ALO') and only the final race of the year (abu-dhabi). By using this argument, I am able to place the label *only* at the very end of the line. Otherwise, ggplot would label each race in the plot. 


::::

<br>
<br>


Here's a plot of the 2022 WDC fight. This year wasn't exactly a nail-biting affair! The championship appears to be over before it even began. In reality, Verstappen mathematically clinched the championship after the Japanese Grand Prix. 


```{r, warning = F, message = F, fig.align='center', fig.height= 10, fig.width=12}


# Pull race results
races2022 <- race_result_scraper(2022)

races2022 %>%
  mutate(rowid = 1:n()) %>%
  group_by(Driver) %>% 
  mutate(cumulativePoints = cumsum(Points)) %>%
  ungroup() %>% 
  ggplot(aes(fct_reorder(Race, rowid), cumulativePoints,
         group = Driver, col = Driver)) +
  geom_line() +
  theme_bw() + 
  theme(
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12),
        axis.text.y = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 18),
        legend.text = element_text(size = 12),
        legend.title = element_text(hjust = 0.5, size = 14)) +
  labs(x = 'Grand Prix',
       y = 'Cumulative Points',
       title = '2022 Formula 1 WDC Race') +
  scale_color_manual(values = c("red", "red", 
                                "cyan3", "cyan3", 
                                "black", "black",
                                "blue", "blue",
                                "darkslateblue", "darkslateblue",
                                "darkred", "darkred", 
                                "darkseagreen4", "darkseagreen4", "darkseagreen4",
                                "darkorange", "darkorange",
                                "cornflowerblue", "cornflowerblue", "cornflowerblue",
                                "navy", "navy"),
                     breaks = c("LEC", "SAI",
                                "HAM", "RUS", 
                                "MAG", "MSC", 
                                "OCO", "ALO",
                                "GAS", "TSU",  
                                "ZHO", "BOT",
                                "STR", "VET", "HUL", 
                                "RIC", "NOR",
                                "LAT", "DEV", "ALB",
                                "PER", "VER"
                                )) +
  geom_text_repel(data = ~ subset(.x, Driver %in% c('VER', 'PER', 'LEC') & Race == 'abu-dhabi'),
            aes(label = Driver),
            nudge_x = 2, nudge_y = 3, show.legend = F)
```



The 2021 Formula 1 season was epic! That year's championship fight was not settled until the final lap of the season. Max Verstappen won the Drivers' Championship for the first time in his career and became the first non-Mercedes driver in the turbo-hybrid era to win the World Championship.

Max Verstappen and Lewis Hamilton engaged in a year-long fight for the title. Verstappen and Hamilton swapped the championship lead multiple times during the 2021 season and entered the season finale, the Abu Dhabi Grand Prix,  *tied on points*. Ultimately, Max overtook Hamilton on the final lap of the Grand Prix after a safety car restart. A truly legendary season. This season also marked the first time since 2008 that the WDC did not race for the WCC (World Constructors Champion). 


```{r, warning = F, message = F, fig.align='center', fig.height= 10, fig.width=12}


# Pull race results
races2021 <- race_result_scraper(2021)

races2021 %>%
  mutate(rowid = 1:n()) %>%
  group_by(Driver) %>% 
  mutate(cumulativePoints = cumsum(Points)) %>%
  ungroup() %>% 
  ggplot(aes(fct_reorder(Race, rowid), cumulativePoints,
         group = Driver, col = Driver)) +
  geom_line() +
  theme_bw() + 
  theme(
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12),
        axis.text.y = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 18),
        legend.text = element_text(size = 12),
        legend.title = element_text(hjust = 0.5, size = 14)) +
  labs(x = 'Grand Prix',
       y = 'Cumulative Points',
       title = '2021 Formula 1 WDC Race') +
  scale_color_manual(values = c("red", "red", 
                                "cyan3", "cyan3", 
                                "black", "black",
                                "blue", "blue",
                                "darkslateblue", "darkslateblue",
                                "darkred", "darkred", "darkred",
                                "darkseagreen4", "darkseagreen4",
                                "darkorange", "darkorange",
                                "cornflowerblue", "cornflowerblue",
                                "navy", "navy"),
                     breaks = c("LEC", "SAI",
                                "HAM", "BOT", 
                                "MAZ", "MSC", 
                                "OCO", "ALO",
                                "GAS", "TSU",  
                                "RAI", "GIO", "KUB",
                                "STR", "VET",  
                                "RIC", "NOR",
                                "LAT", "RUS",
                                "PER", "VER"
                                )) +
  geom_text_repel(data = ~ subset(.x, Driver %in% c('VER', 'HAM', 'BOT', 'PER') & Race == 'abu-dhabi'),
            aes(label = Driver),
            nudge_x = 1.5, nudge_y = 3, show.legend = F)
```



The 2020 season was postponed until July due to the COVID-19 pandemic. A total of 17 races took place in 2020 because the schedule experienced several cancellations and new replacement races. Lewis Hamilton and Mercedes continued their dominance and won their seventh WDC and WCC, respectively. Mercedes locked up the WCC at the Emilia Romagna Grand Prix, while Lewis Hamilton tied Michael Schumacher's record of seven WDC.


```{r, warning = F, message = F, fig.align='center', fig.height= 10, fig.width=12}


# Pull race results
races2020 <- race_result_scraper(2020)

races2020 %>%
  mutate(rowid = 1:n()) %>%
  group_by(Driver) %>% 
  mutate(cumulativePoints = cumsum(Points)) %>%
  ungroup() %>% 
  ggplot(aes(fct_reorder(Race, rowid), cumulativePoints,
         group = Driver, col = Driver)) +
  geom_line() +
  theme_bw() + 
  theme(
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12),
        axis.text.y = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 18),
        legend.text = element_text(size = 12),
        legend.title = element_text(hjust = 0.5, size = 14)) +
  labs(x = 'Grand Prix',
       y = 'Cumulative Points',
       title = '2020 Formula 1 WDC Race') +
  geom_text_repel(data = ~ subset(.x, Driver %in% c('VER', 'HAM', 'BOT', 'PER') & Race == 'abu-dhabi'),
            aes(label = Driver),
            nudge_x = 1.5, nudge_y = 3, show.legend = F) +
  scale_color_manual(values = c("red", "red", 
                                "cyan3", "cyan3", 
                                "black", "black", "black",
                                "yellow", "yellow",
                                "darkslateblue", "darkslateblue",
                                "darkred", "darkred", 
                                "pink", "pink", "pink",
                                "darkorange", "darkorange",
                                "cornflowerblue", "cornflowerblue", "cornflowerblue",
                                "navy", "navy"
                                ),
                     breaks = c("LEC", "VET",
                                "HAM", "BOT",
                                "MAG", "GRO", "FIT",
                                "OCO", "RIC",
                                "GAS", "KVY",  
                                "RAI", "GIO",
                                "STR", "PER", "HUL",  
                                "SAI", "NOR",
                                "LAT", "RUS", "AIT",
                                "ALB", "VER"
                                )) 
```



In 2019, Lewis Hamilton and Mercedes both defended their championships, winning their sixth. Max Verstappen was able to finish third, ahead of both Ferrari drivers, in the WDC. 


```{r, warning = F, message = F, fig.align='center', fig.height= 10, fig.width=12}


# Pull race results
races2019 <- race_result_scraper(2019)

races2019 %>%
  mutate(rowid = 1:n()) %>%
  group_by(Driver) %>% 
  mutate(cumulativePoints = cumsum(Points)) %>%
  ungroup() %>% 
  ggplot(aes(fct_reorder(Race, rowid), cumulativePoints,
         group = Driver, col = Driver)) +
  geom_line() +
  theme_bw() + 
  theme(
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12),
        axis.text.y = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 18),
        legend.text = element_text(size = 12),
        legend.title = element_text(hjust = 0.5, size = 14)) +
  labs(x = 'Grand Prix',
       y = 'Cumulative Points',
       title = '2019 Formula 1 WDC Race') +
  geom_text_repel(data = ~ subset(.x, Driver %in% c('VER', 'HAM', 'BOT', 'LEC', 'VET') & Race == 'abu-dhabi'),
            aes(label = Driver),
            nudge_x = 1.5, nudge_y = 1.5, show.legend = F) +
  scale_color_manual(values = c("red", "red", 
                                "cyan3", "cyan3", 
                                "black", "black", 
                                "yellow", "yellow",
                                "darkslateblue",
                                "darkred", "darkred", 
                                "pink", "pink", "pink",
                                "darkorange", "darkorange",
                                "cornflowerblue", "cornflowerblue",
                                "navy", "navy", "navy"),
                     breaks = c("LEC", "VET",
                                "HAM", "BOT",
                                "MAG", "GRO",
                                "OCO", "RIC",
                                "KVY",
                                "RAI", "GIO",
                                "STR", "PER",  
                                "SAI", "NOR",
                                "LAT", "RUS",
                                "ALB", "VER", "GAS"
                                )) 
```



For the second straight year, Mercedes and Ferrari battled for the WCC. Two four-time WDCs, Lewis Hamilton and Sebastian Vettel, battled for a fifth championship. Hamilton and Vettel swapped the WDC lead five times throughout 2018. In fact, Vettel held the lead at the midway point of the season. But, in the end, Hamilton prevailed and secured his fifth title. 


```{r, warning = F, message = F, fig.align='center', fig.height= 10, fig.width=12}


# Pull race results
races2018 <- race_result_scraper(2018)

races2018 %>%
  mutate(rowid = 1:n()) %>%
  group_by(Driver) %>% 
  mutate(cumulativePoints = cumsum(Points)) %>%
  ungroup() %>% 
  ggplot(aes(fct_reorder(Race, rowid), cumulativePoints,
         group = Driver, col = Driver)) +
  geom_line() +
  theme_bw() + 
  theme(
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12),
        axis.text.y = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 18),
        legend.text = element_text(size = 12),
        legend.title = element_text(hjust = 0.5, size = 14)) +
  labs(x = 'Grand Prix',
       y = 'Cumulative Points',
       title = '2018 Formula 1 WDC Race') +
  geom_text_repel(data = ~ subset(.x, Driver %in% c('VER', 'HAM', 'BOT', 'RAI', 'VET', 'RIC') & Race == 'abu-dhabi'),
            aes(label = Driver),
            nudge_x = 1.5, nudge_y = 1, show.legend = F) + 
  scale_color_manual(values = c("red", "red", 
                                "cyan3", "cyan3", 
                                "black", "black", 
                                "yellow", "yellow",
                                "darkslateblue", "darkslateblue",
                                "darkred", "darkred", 
                                "pink", "pink",
                                "darkorange", "darkorange",
                                "cornflowerblue", "cornflowerblue",
                                "navy", "navy"),
                     breaks = c("RAI", "VET",
                                "HAM", "BOT",
                                "MAG", "GRO",
                                "SAI", "HUL",
                                "HAR", "GAS",
                                "ERI", "LEC",
                                "OCO", "PER",  
                                "VAN", "ALO",
                                "STR", "SIR",
                                "RIC", "VER"
                                )) 
```


After three straight years of teammates battling for the WDC, there was a legitimate battle *between* teams in 2017. Lewis Hamilton battled four-time WDC Sebastian Vettel. Vettel lead the WDC for the first 12 rounds of the season, but Hamilton ultimately won his fourth championship by 46 points. 


```{r, warning = F, message = F, fig.align='center', fig.height= 10, fig.width=12}


# Pull race results
races2017 <- race_result_scraper(2017)

races2017 %>%
  mutate(rowid = 1:n()) %>%
  group_by(Driver) %>% 
  mutate(cumulativePoints = cumsum(Points)) %>%
  ungroup() %>% 
  ggplot(aes(fct_reorder(Race, rowid), cumulativePoints,
         group = Driver, col = Driver)) +
  geom_line() +
  theme_bw() + 
  theme(
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12),
        axis.text.y = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 18),
        legend.text = element_text(size = 12),
        legend.title = element_text(hjust = 0.5, size = 14)) +
  labs(x = 'Grand Prix',
       y = 'Cumulative Points',
       title = '2017 Formula 1 WDC Race') + geom_text_repel(data = ~ subset(.x, Driver %in% c('RIC', 'HAM', 'BOT', 'RAI', 'VET', 'VER') & Race == 'abu-dhabi'),
            aes(label = Driver),
            nudge_x = 1.5, nudge_y = 1, show.legend = F) + 
  scale_color_manual(values = c("red", "red", 
                                "cyan3", "cyan3", 
                                "black", "black", 
                                "yellow", "yellow", "yellow",
                                "darkslateblue", "darkslateblue", "darkslateblue", 
                                "darkred", "darkred", "darkred", 
                                "pink", "pink", 
                                "darkorange", "darkorange", "darkorange",
                                "cornflowerblue", "cornflowerblue", "cornflowerblue",
                                "navy", "navy"),
                     breaks = c("RAI", "VET",
                                "HAM", "BOT",
                                "MAG", "GRO",
                                "SAI", "HUL", "PAL",
                                "HAR", "GAS", "KYV",
                                "ERI", "GIO", "WEH",
                                "OCO", "PER",  
                                "VAN", "ALO", "BUT",
                                "STR", "MAS", "DIR",
                                "RIC", "VER"
                                )) 
```


In 2016, Nico Rosberg won his first and only WDC title in the final race of the season. Rosberg beat his teammate Lewis Hamilton thanks to nine race wins and seven additional podiums. With this WDC title, Nico became only the second son of a champion to become champion himself. Pretty cool! Following this title win, Nico retired from Formula 1. 



```{r, warning = F, message = F, fig.align='center', fig.height= 10, fig.width=12}


# Pull race results
races2016 <- race_result_scraper(2016)

races2016 %>%
  mutate(rowid = 1:n()) %>%
  group_by(Driver) %>% 
  mutate(cumulativePoints = cumsum(Points)) %>%
  ungroup() %>% 
  ggplot(aes(fct_reorder(Race, rowid), cumulativePoints,
         group = Driver, col = Driver)) +
  geom_line() +
  theme_bw() + 
  theme(
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12),
        axis.text.y = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 18),
        legend.text = element_text(size = 12),
        legend.title = element_text(hjust = 0.5, size = 14)) +
  labs(x = 'Grand Prix',
       y = 'Cumulative Points',
       title = '2016 Formula 1 WDC Race') + geom_text_repel(data = ~ subset(.x, Driver %in% c('RIC', 'HAM', 'ROS', 'RAI', 'VET', 'VER') & Race == 'abu-dhabi'),
            aes(label = Driver),
            nudge_x = 1.5, nudge_y = 1, show.legend = F) + 
  scale_color_manual(values = c("red", "red", 
                                "cyan3", "cyan3", 
                                "black", "black", 
                                "yellow", "yellow", 
                                "darkslateblue", "darkslateblue",
                                "darkred", "darkred", 
                                "pink", "pink", 
                                "darkorange", "darkorange", "darkorange",
                                "cornflowerblue", "cornflowerblue", 
                                "navy", "navy",
                                "blue", "blue", "blue"),
                     breaks = c("RAI", "VET",
                                "HAM", "ROS",
                                "GUT", "GRO",
                                "MAG", "PAL", 
                                "SAI", "KYV",
                                "ERI", "NAS", 
                                "HUL", "PER",  
                                "VAN", "ALO", "BUT",
                                "BOT", "MAS",
                                "RIC", "VER",
                                "HAR", "OCO", "WEH"
                                )) 
```



Lewis Hamilton won his third overall and second straight WDC in 2015. He managed to beat his teammate Nico Rosberg and four-time WDC Sebastian Vettel. 

```{r, warning = F, message = F, fig.align='center', fig.height= 10, fig.width=12}


# Pull race results
races2015 <- race_result_scraper(2015)

races2015 %>%
  mutate(rowid = 1:n()) %>%
  group_by(Driver) %>% 
  mutate(cumulativePoints = cumsum(Points)) %>%
  ungroup() %>% 
  ggplot(aes(fct_reorder(Race, rowid), cumulativePoints,
         group = Driver, col = Driver)) +
  geom_line() +
  theme_bw() + 
  theme(
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12),
        axis.text.y = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 18),
        legend.text = element_text(size = 12),
        legend.title = element_text(hjust = 0.5, size = 14)) +
  labs(x = 'Grand Prix',
       y = 'Cumulative Points',
       title = '2015 Formula 1 WDC Race') + geom_text_repel(data = ~ subset(.x, Driver %in% c('HAM', 'ROS', 'RAI', 'VET') & Race == 'abu-dhabi'),
            aes(label = Driver),
            nudge_x = 1.5, nudge_y = 1, show.legend = F) + 
  scale_color_manual(values = c("red", "red", 
                                "cyan3", "cyan3", 
                                "black", "black", 
                                "grey", "grey", "grey",
                                "darkslateblue", "darkslateblue",
                                "darkred", "darkred", 
                                "pink", "pink", 
                                "darkorange", "darkorange", "darkorange",
                                "cornflowerblue", "cornflowerblue", 
                                "navy", "navy",
                                "blue", "blue", "blue"),
                     breaks = c("RAI", "VET",
                                "HAM", "ROS",
                                "MAL", "GRO",
                                "STE", "WEH", "ROS",
                                "SAI", "VER",
                                "ERI", "NAS", 
                                "HUL", "PER",  
                                "MAG", "ALO", "BUT",
                                "BOT", "MAS",
                                "RIC", "KYV",
                                "HAR", "OCO", "WEH"
                                )) 
```


2014 was an important year in Formula 1. It marked the beginning of a Mercedes dynasty, the onset of a Lewis Hamilton reign, and the ushering in of a revised engine formula. In 2014, the previous era's V8 engine was replaced with a turbocharged V6 engine that included an energy recovery system (ERS).

Sebastian Vettel and Red Bull were the defending WDC and WCC, respectively. However, it became immediately evident that Mercedes would dominate this season. Lewis Hamilton won his second WDC (his previous title was with McLaren) and Mercedes dominated the WCC fight, beating Red Bull by nearly a 300 point margin.


```{r, warning = F, message = F, fig.align='center', fig.height= 10, fig.width=12}


# Pull race results
races2014 <- race_result_scraper(2014)

races2014 %>%
  mutate(rowid = 1:n()) %>%
  group_by(Driver) %>% 
  mutate(cumulativePoints = cumsum(Points)) %>%
  ungroup() %>% 
  ggplot(aes(fct_reorder(Race, rowid), cumulativePoints,
         group = Driver, col = Driver)) +
  geom_line() +
  theme_bw() + 
  theme(
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 12),
        axis.text.y = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 18),
        legend.text = element_text(size = 12),
        legend.title = element_text(hjust = 0.5, size = 14)) +
  labs(x = 'Grand Prix',
       y = 'Cumulative Points',
       title = '2014 Formula 1 WDC Race') + geom_text_repel(data = ~ subset(.x, Driver %in% c('HAM', 'ROS', 'RIC', 'VET') & Race == 'abu-dhabi'),
            aes(label = Driver),
            nudge_x = 1.5, nudge_y = 1, show.legend = F) + 
  scale_color_manual(values = c("red", "red", 
                                "cyan3", "cyan3", 
                                "black", "black", 
                                "grey", "grey", "grey",
                                "darkslateblue", "darkslateblue",
                                "darkred", "darkred", 
                                "pink", "pink", 
                                "darkorange", "darkorange", 
                                "cornflowerblue", "cornflowerblue", 
                                "navy", "navy",
                                "seagreen", "seagreen", "seagreen",  "seagreen"),
                     breaks = c("RAI", "ALO",
                                "HAM", "ROS",
                                "MAL", "GRO",
                                "CHI", "BIA", "ROS",
                                "KYV", "VER",
                                "SUT", "GUT", 
                                "HUL", "PER",  
                                "MAG",  "BUT",
                                "BOT", "MAS",
                                "RIC", "VET",
                                "ERI", "STE", "KOB", "LOT"
                                )) 
```



\newpage


